### 2023-01-26 paper-thesis/chapters/000-abstract.tex
Cleanup work

Static type systems can greatly enhance the quality of programs, but implementing a type checker that is both expressive and user-friendly is challenging and error-prone. The Statix meta-language (part of the Spoofax language workbench) aims to make this task easier by automatically deriving a type checker from a declarative specification of the type system. However, so far Statix has not been used to implement dependent types, an expressive class of type systems which require evaluation of terms during type checking. In this paper, we present an implementation of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments. While we encountered some challenges in the implementation, our conclusion is that Statix is already usable as a tool for implementing dependent types.

>>>>

Static type systems can greatly enhance the quality of programs, but implementing a type checker that is both expressive and user-friendly is challenging and error-prone. The Statix meta-language (part of the Spoofax language workbench) aims to make this task easier by automatically deriving a type checker from a declarative specification of the type system. However, so far Statix has not been used to implement dependent types, an expressive class of type systems which require evaluation of terms during type checking. In this thesis, we present an implementation of a simple dependently typed language in Statix, and discuss how to extend it with several common features such as inductive data types, universes, and inference of implicit arguments. While we encountered some challenges in the implementation, our conclusion is that Statix is already usable as a tool for implementing dependent types.

### 2023-01-26 paper-thesis/chapters/01-introduction.tex
Cleanup work

Spoofax is a textual language workbench: \label{key}a collection of tools that enable the development of textual languages \cite{spoofax}. When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics. To provide these advantages to as many language developers as possible, Statix aims to cover a broad range of languages and type systems. However, no attempts have been made to express dependently typed languages in Statix.

Dependently typed languages are different from other languages, because they allow types to be parameterized by values. This allows types to express properties of values that cannot be expressed in a simple type system, such as the length of a list or the well-formedness of a binary search tree. This expressiveness also makes dependent type systems more complicated to implement. Especially, deciding equality of types requires evaluation of the terms they are parameterized by.

\section{Contributions}
This goal of this paper is to investigate how well Statix is fit for the task of defining a simple dependently-typed language. We want to investigate whether typical features of dependently typed languages can be encoded concisely in Statix. The goal is not only to show that Statix can implement it, but to investigate whether implementing a dependent type checker is easier in Statix than in a general-purpose programming language.

The technical contributions of this thesis are the following:
\begin{itemize}
	\item 
\end{itemize}

\section{Outline}

TODO navigation

>>>>

Spoofax is a language workbench: a collection of tools that enable the development of textual languages~\cite{spoofax}. When working with the Spoofax workbench, the Statix meta-language can be used for the specification of static semantics. It is a declarative language that uses inference rules to define static semantics. From these rules a type checker and other editor tools can be derived. Statix aims to cover a broad range of languages and type systems. However, no attempts have been made to express dependently typed languages in Statix until now.

Dependently typed languages are different from other languages, because they allow types to be parameterized by values. This allows types to express properties of values that cannot be expressed in a simple type system, such as the length of a list or the well-formedness of a binary search tree. This expressiveness also makes dependent type systems more complicated to implement. Especially, deciding equality of types requires normalization of the terms they are parameterized by.

\section*{Contributions}
This goal of this thesis is to investigate how well Statix is fit for the task of defining a simple dependently-typed language. We want to investigate whether typical features of dependently typed languages can be encoded concisely in Statix. The goal is not only to show that Statix can implement it, but also that the implementation is concise.

The technical contributions of this thesis are the following:
\begin{itemize}
	\item We present a implementation of the Calculus of Constructions in Statix (Chapter \ref{chap:baselang})
	\item We show the implementation is easily extendable (Chapter \ref{chap:bools})
	\item We show how to add inference to the implementation (Chapter \ref{ch:inference})
	\item We show how to add support for inductive datatypes to the implementation (Chapter \ref{ch:datatypes})
	\item We show how to add support for universes to the implementation (Chapter \ref{ch:universes})
\end{itemize}

\section*{Discussions}

Furthermore, the thesis contains some discussions:
\begin{itemize}
	\item We discuss how well semantic code completion works for our language (Chapter \ref{chap:editor-services})
	\item We compare our implementation with an implementation of the same language in Haskell (Chapter \ref{ch:comp-haskell})
	\item We compare our implementation with an implementation of the same language in LambdaPi (Chapter \ref{ch:comp-lambdapi})
	\item We discuss how Spoofax can be improved to better support implementing dependently typed languages (Chapter \ref{ch:ergonomics})
	\item We discuss related work (Chapter \ref{ch:relatedwork}) and future work (Chapter \ref{ch:futurework})

\end{itemize}

### 2023-01-26 paper-thesis/chapters/02-background-spoofax.tex
Cleanup work

The Spoofax Language Workbench \cite{spoofax} is a platform to develop domain-specific and
general-purpose programming languages. Each aspect of the language is defined in one of three metalanguages, each with their own purpose. From these languages Spoofax generates a parser, type checker and other useful tools.

The metalanguage SDF3 is used to define the syntax of the language \cite{sdf3}, which is then parsed using Spoofax's SGLR parser \cite{sdf3_parser}. The parser produces a abstract syntax tree in the ATerm format \cite{aterm}. A pretty printer is also generated based on this syntax definition.

Next, the metalanguage Statix \cite{scopes_as_types} is used to define the static semantics of the language. It uses scope graphs to represent binding and typing information of the language. How this works is discussed in section \ref{sec:statix}.

Finally, Stratego \cite{stratego} is used to define program transformations, using rewrite rules. This can be used for modifications to the AST, but also to build transformations to a different language (such as compiling to assembly).

@@

\textit{Note: This section is adapted from the Explanation in "Composable Type System Specification
	using Heterogeneous Scope Graphs". \cite[sect. 4.1.2]{Zwaan21}}

In Statix, scope graphs are used to represent a programs name and type information \cite{scopes_as_types}. A scope graph is a graph that consists of the following components:

>>>>

The Spoofax Language Workbench~\cite{spoofax} is a platform to develop domain-specific and
general-purpose programming languages. Each aspect of the language is defined in one of three metalanguages, each with their own purpose. From these languages Spoofax generates a parser, type checker and other useful tools.

The metalanguage SDF3 is used to define the syntax of the language~\cite{sdf3}, which is then parsed using Spoofax's SGLR parser~\cite{sdf3_parser}. The parser produces a abstract syntax tree in the ATerm format~\cite{aterm}. A pretty printer is also generated based on this syntax definition.

Next, the metalanguage Statix~\cite{scopes_as_types} is used to define the static semantics of the language. It uses scope graphs to represent binding and typing information of the language. How this works is discussed in section \ref{sec:statix}.

Finally, Stratego~\cite{stratego} is used to define program transformations, using rewrite rules. This can be used for modifications to the AST, but also to build transformations to a different language (such as compiling to assembly).

@@

\textit{Note: This section is adapted from the Explanation in "Composable Type System Specification
	using Heterogeneous Scope Graphs".~\cite[sect. 4.1.2]{Zwaan21}}

In Statix, scope graphs are used to represent a programs name and type information~\cite{scopes_as_types}. A scope graph is a graph that consists of the following components:

### 2023-01-26 paper-thesis/chapters/03-background-dependent-types.tex
Update 03-background-dependent-types.tex
Cleanup work

The Curry-Howard correspondence is a relationship that can be used to interpret typed programs as mathematical proofs \cite{chc}. This is done by representing false statements as empty types, and true statements as non-empty (inhabited) types.

@@

The lambda cube is a framework to categorize programming languages based on whether terms and types can depend on each other \cite{lambda_cube}. Specifically, it categorizes languages on three axis: \footnote{Note that the "Terms can depend on terms" axis is missing, this is because a language where terms cannot depend on terms cannot compute anything and is thus pretty useless.}

>>>>

The Curry-Howard correspondence is a relationship that can be used to interpret typed programs as mathematical proofs~\cite{chc}. This is done by representing false statements as empty types, and true statements as non-empty (inhabited) types.

@@

The lambda cube is a framework to categorize programming languages based on whether terms and types can depend on each other~\cite{lambda_cube}. Specifically, it categorizes languages on three axis: \footnote{Note that the "Terms can depend on terms" axis is missing, this is because a language where terms cannot depend on terms cannot compute anything and is thus pretty useless.}
